---
title: "BU52710 Predictive Analytics – Checkpoint Submission"
author: 'Andrew Wilson Student ID: 2417549'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

# Introduction

This checkpoint report builds a predictive model to identify customers of Broughty Ferry Media that are most likely to renew their magazine subscriptions. The business objective is to maximise profit from a £20,000 direct-mail campaign targeting a maximum of 500 customers (£40 per contact, £140 revenue per renewal).

---

# Step 1: Data Loading and Train/Validation Split

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(caret)
library(pROC)
library(rpart)
library(rpart.plot)
library(randomForest)
library(car)       # for VIF
library(knitr)
library(kableExtra)
library(ggplot2)
library(corrplot)
library(gridExtra)
```

```{r load-data}
# Load the dataset
data <- read.csv("data__2417549.csv", stringsAsFactors = TRUE)

# Convert target variable to factor
data$Renewal <- factor(data$Renewal, levels = c("No", "Yes"))

# Quick overview
dim(data)
str(data)
```

```{r train-val-split}
# Reproducible split: 70% training, 30% validation
set.seed(2417549)
split_index <- createDataPartition(data$Renewal, p = 0.70, list = FALSE)

train_data      <- data[split_index, ]
validation_data <- data[-split_index, ]

cat("Training set:  ", nrow(train_data), "rows\n")
cat("Validation set:", nrow(validation_data), "rows\n")

# Verify the class balance is preserved
train_rate <- round(prop.table(table(train_data$Renewal)) * 100, 2)
val_rate   <- round(prop.table(table(validation_data$Renewal)) * 100, 2)

balance_check <- data.frame(
  Set        = c("Training", "Validation"),
  No_pct     = c(train_rate["No"],  val_rate["No"]),
  Yes_pct    = c(train_rate["Yes"], val_rate["Yes"])
)
kable(balance_check,
      col.names = c("Set", "% Non-Renewal", "% Renewal"),
      caption   = "Renewal Rate Balance Across Splits",
      digits    = 2) %>%
  kable_styling(full_width = FALSE)
```

*Stratified sampling via `createDataPartition` ensures that both sets have similar renewal rates, preventing any data leakage and preserving the representativeness of the data.*

---

# Step 2: Exploratory Data Analysis

## Overall Renewal Rate

```{r renewal-rate}
renewal_summary <- data %>%
  count(Renewal) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2))

kable(renewal_summary,
      col.names = c("Renewal", "Count", "Percentage (%)"),
      caption   = "Overall Renewal Distribution") %>%
  kable_styling(full_width = FALSE)
```

## Check for Missing Values

```{r missing-values}
missing_vals <- data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  filter(Missing_Count > 0)

if (nrow(missing_vals) == 0) {
  cat("No missing values detected in the dataset.\n")
} else {
  kable(missing_vals, caption = "Variables with Missing Values") %>%
    kable_styling(full_width = FALSE)
}
```

## Exploratory Analysis - Renewal Rate by Potential Key Variables

```{r eda-plots, fig.height=5}
# Plot 1: Renewal by Active Subscriptions
p1 <- ggplot(data, aes(x = factor(ActiveSubscriptions), fill = Renewal)) +
  geom_bar(position = "fill") +
  labs(title = "Renewal Rate by Active Subscriptions",
       x = "Active Subscriptions", y = "Proportion") +
  scale_fill_manual(values = c("#C2C2C2", "#1E3A8A")) +
  theme_minimal()

# Plot 2: Total Amount Paid distribution by renewal
p2 <- ggplot(data, aes(x = Renewal, y = TotalAmountPaid, fill = Renewal)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Total Amount Paid by Renewal Status",
       x = "Renewal", y = "Total Amount Paid (£)") +
  scale_fill_manual(values = c("#C2C2C2", "#1E3A8A")) +
  theme_minimal() +
  theme(legend.position = "none")

# Plot 3: Years Since Last Order
p3 <- ggplot(data, aes(x = Renewal, y = YearsSinceLastOrder, fill = Renewal)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Years Since Last Order by Renewal",
       x = "Renewal", y = "Years Since Last Order") +
  scale_fill_manual(values = c("#C2C2C2", "#1E3A8A")) +
  theme_minimal() +
  theme(legend.position = "none")

# Plot 4: Paid Direct Mail Orders
p4 <- ggplot(data, aes(x = Renewal, y = PaidDirectMailOrders, fill = Renewal)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Paid Direct Mail Orders by Renewal",
       x = "Renewal", y = "Paid Direct Mail Orders") +
  scale_fill_manual(values = c("#C2C2C2", "#1E3A8A")) +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Numeric Variable Summary

```{r summary-table}
numeric_vars <- data %>% select(where(is.numeric)) %>% select(-CustomerID)
summary_stats <- numeric_vars %>%
  pivot_longer(everything(), names_to = "Variable") %>%
  group_by(Variable) %>%
  summarise(
    Mean   = round(mean(value, na.rm = TRUE), 2),
    SD     = round(sd(value, na.rm = TRUE), 2),
    Min    = round(min(value, na.rm = TRUE), 2),
    Max    = round(max(value, na.rm = TRUE), 2)
  )

kable(summary_stats, caption = "Summary Statistics for Numeric Variables") %>%
  kable_styling(full_width = FALSE, font_size = 9)
```

---

# Step 3: Variable Selection (10 Variables)

## Framework for Variable Selection

The 10 variables should be chosen based on a combination of:
- Theoretical relevance to subscription renewal behaviour (RMF)
- Correlation analysis with the target variable
- Avoidance of multicollinearity (VIF check)
- Business interpretability

```{r correlation-all-vars}
# Point-biserial correlation of ALL numeric variables with Renewal
renewal_numeric <- as.numeric(train_data$Renewal) - 1

# Select all numeric variables excluding CustomerID
all_numeric_vars <- train_data %>%
  select(where(is.numeric)) %>%
  select(-CustomerID) %>%
  names()

# Calculate point-biserial correlation for every numeric variable
all_corr <- sapply(all_numeric_vars, function(v) {
  cor(train_data[[v]], renewal_numeric, use = "complete.obs")
})

# Build and display results table, sorted by absolute correlation
all_corr_df <- data.frame(
  Variable    = names(all_corr),
  Correlation = round(all_corr, 4)
) %>% arrange(desc(abs(Correlation)))

kable(all_corr_df,
      col.names = c("Variable", "Point-Biserial Correlation with Renewal"),
      caption   = "Correlation of All Numeric Variables with Renewal",
      row.names = FALSE) %>%
  kable_styling(full_width = FALSE)
```

**Selected Variables:**

| Variable | Type | Rationale |
|---|---|---|
| `PoundsPerIssue` | Numeric | Strongest predictor; spend per issue reflects engagement value |
| `Income` | Numeric | Financial capacity to renew; higher income customers are less price-sensitive |
| `Child6.12` | Numeric | Household composition influences subscription habits |
| `TotalAmountPaid` | Numeric | Cumulative monetary spend signals long-term engagement and loyalty |
| `MonthsSince1stOrder` | Numeric | Customer tenure; longer relationships often correlate with higher loyalty |
| `TotalPaidOrders` | Numeric | Order frequency reflects habitual purchasing behaviour |
| `Child13.18` | Numeric | Household composition; older children may drive magazine interest |
| `PaidCashMagazines` | Numeric | Cash payment behaviour indicates active and committed engagement |
| `HouseholdSize` | Numeric | Larger households may have multiple readers, supporting renewal |
| `HomeValue` | Numeric | Proxy for financial stability and disposable income |

```{r variable-selection}
# Define the selected variables
selected_vars <- c("PoundsPerIssue", "Income", "Child6.12",
                   "TotalAmountPaid", "MonthsSince1stOrder", "TotalPaidOrders",
                   "Child13.18", "PaidCashMagazines",
                   "HouseholdSize", "HomeValue")

# Creation of Train and Validation Subsets
train_sel <- train_data %>% select(all_of(selected_vars), Renewal)
val_sel   <- validation_data %>% select(all_of(selected_vars), Renewal)
```

## Multicollinearity – Variance Inflation Factors (VIF) Test

```{r vif-check}
# Fit a logistic regression with the selected variables to calculate VIF
vif_model <- glm(Renewal ~ ., data = train_sel, family = binomial)
vif_values <- vif(vif_model)

vif_df <- data.frame(
  Variable = names(vif_values),
  VIF      = round(vif_values, 3)
)

kable(vif_df,
      col.names = c("Variable", "VIF"),
      caption   = "Variance Inflation Factors for Selected Variables",
      row.names = FALSE) %>%
  kable_styling(full_width = FALSE)

cat("\nAll VIF values below 10 indicate no severe multicollinearity.\n")
```

---

# Step 4: Build Three Predictive Models

## Model 1: Logistic Regression

```{r logistic-regression}
# Train logistic regression
logit_model <- glm(Renewal ~ ., data = train_sel, family = binomial(link = "logit"))
summary(logit_model)

# Predict on validation set
logit_probs <- predict(logit_model, newdata = val_sel, type = "response")
logit_pred  <- ifelse(logit_probs >= 0.5, "Yes", "No")
logit_pred  <- factor(logit_pred, levels = c("No", "Yes"))

# Confusion Matrix
cm_logit <- confusionMatrix(logit_pred, val_sel$Renewal, positive = "Yes")
print(cm_logit)

# Receiver Operating Characteristic (ROC) and Area Under Curve (AUC)
roc_logit <- roc(as.numeric(val_sel$Renewal) - 1, logit_probs)
cat("\nLogistic Regression AUC:", round(auc(roc_logit), 4), "\n")
```

## Model 2: Decision Tree

```{r decision-tree}
# Train decision tree
tree_model <- rpart(Renewal ~ ., data = train_sel,
                    method = "class",
                    control = rpart.control(cp = 0.001, maxdepth = 6))

# Plot tree
rpart.plot(tree_model, type = 3, extra = 104, fallen.leaves = TRUE,
           main = "Decision Tree – Customer Renewal")

# Predict on validation set
tree_probs <- predict(tree_model, newdata = val_sel, type = "prob")[, "Yes"]
tree_pred  <- ifelse(tree_probs >= 0.5, "Yes", "No")
tree_pred  <- factor(tree_pred, levels = c("No", "Yes"))

# Confusion Matrix
cm_tree <- confusionMatrix(tree_pred, val_sel$Renewal, positive = "Yes")
print(cm_tree)

# Receiver Operating Characteristic (ROC) and Area Under Curve (AUC)
roc_tree <- roc(as.numeric(val_sel$Renewal) - 1, tree_probs)
cat("\nDecision Tree AUC:", round(auc(roc_tree), 4), "\n")
```

## Model 3: Random Forest

```{r random-forest}
set.seed(2417549) # For reproducibility
# Train random forest
rf_model <- randomForest(Renewal ~ ., data = train_sel,
                         ntree = 500, mtry = 3,
                         importance = TRUE)

print(rf_model)

# Predict on validation set
rf_probs <- predict(rf_model, newdata = val_sel, type = "prob")[, "Yes"]
rf_pred  <- predict(rf_model, newdata = val_sel, type = "class")

# Confusion Matrix
cm_rf <- confusionMatrix(rf_pred, val_sel$Renewal, positive = "Yes")
print(cm_rf)

# Receiver Operating Characteristic (ROC) and Area Under Curve (AUC)
roc_rf <- roc(as.numeric(val_sel$Renewal) - 1, rf_probs)
cat("\nRandom Forest AUC:", round(auc(roc_rf), 4), "\n")

# Variable Importance
importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

ggplot(importance_df, aes(x = reorder(Variable, MeanDecreaseGini),
                           y = MeanDecreaseGini)) +
  geom_col(fill = "#1E3A8A", alpha = 0.85) +
  coord_flip() +
  labs(title    = "Random Forest – Variable Importance",
       subtitle = "Measured by Mean Decrease in Gini Impurity",
       x        = NULL,
       y        = "Mean Decrease Gini") +
  theme_minimal() +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(size = 10, colour = "grey40"),
    axis.text          = element_text(size = 10, colour = "grey20"),
    axis.title.x       = element_text(size = 11, colour = "grey20"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(colour = "grey85")
  )
```

---

# Step 5: Model Comparison and Selection

## Performance Comparison Table

```{r model-comparison}
model_comparison <- data.frame(
  Model    = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = round(c(cm_logit$overall["Accuracy"],
                     cm_tree$overall["Accuracy"],
                     cm_rf$overall["Accuracy"]), 4),
  Sensitivity = round(c(cm_logit$byClass["Sensitivity"],
                        cm_tree$byClass["Sensitivity"],
                        cm_rf$byClass["Sensitivity"]), 4),
  Specificity = round(c(cm_logit$byClass["Specificity"],
                        cm_tree$byClass["Specificity"],
                        cm_rf$byClass["Specificity"]), 4),
  Precision = round(c(cm_logit$byClass["Pos Pred Value"],
                      cm_tree$byClass["Pos Pred Value"],
                      cm_rf$byClass["Pos Pred Value"]), 4),
  AUC      = round(c(auc(roc_logit), auc(roc_tree), auc(roc_rf)), 4)
)

kable(model_comparison,
      col.names = c("Model", "Accuracy", "Sensitivity", "Specificity", "Precision", "AUC"),
      caption   = "Model Performance on Validation Set") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(which.max(model_comparison$AUC), bold = TRUE, background = "#D5F5E3")
```

## Receiver Operating Characteristic (ROC) Curve Comparison

```{r roc-curves}
plot(roc_logit, col = "#A0BDD8", lwd = 2,
     main = "ROC Curves – All Models (Validation Set)")
plot(roc_tree, col = "#6B9DC2", lwd = 2, add = TRUE)
plot(roc_rf,   col = "#1E3A8A", lwd = 2, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey50")
legend("bottomright",
       legend = c(paste("Logistic Regression (AUC =", round(auc(roc_logit), 3), ")"),
                  paste("Decision Tree (AUC =",       round(auc(roc_tree), 3), ")"),
                  paste("Random Forest (AUC =",       round(auc(roc_rf), 3), ")")),
       col    = c("#A0BDD8", "#6B9DC2", "#1E3A8A"),
       lwd    = c(2, 2, 3))
```

## Profit Simulation on Validation Set (Top 500 Equivalent)

```{r profit-simulation}
# Business constraints
cost_per_contact <- 40
revenue_per_renewal <- 140
budget <- 20000
max_contacts <- budget / cost_per_contact  # 500 at full scale

# Scale top-n to validation set (30% of 8000 = 2400 rows)
n_val <- nrow(validation_data) / nrow(data)
top_n_val <- round(max_contacts * n_val)  # proportional top-n for validation

# Random baseline: expected renewals from random selection
set.seed(2417549)
baseline_rate <- mean(val_sel$Renewal == "Yes")
random_renewals <- round(top_n_val * baseline_rate)
random_profit <- random_renewals * revenue_per_renewal - top_n_val * cost_per_contact

# Function to calculate the profit given ranked probabilities
calculate_profit <- function(probs, actual, top_n, cost, rev) {
  ranked <- order(probs, decreasing = TRUE)[1:top_n]
  renewals <- sum(actual[ranked] == "Yes")
  profit <- renewals * rev - top_n * cost
  return(list(renewals = renewals, profit = profit, top_n = top_n))
}

# Calculate Profit for each model
prof_logit <- calculate_profit(logit_probs, val_sel$Renewal, top_n_val, cost_per_contact, revenue_per_renewal)
prof_tree  <- calculate_profit(tree_probs,  val_sel$Renewal, top_n_val, cost_per_contact, revenue_per_renewal)
prof_rf    <- calculate_profit(rf_probs,    val_sel$Renewal, top_n_val, cost_per_contact, revenue_per_renewal)

profit_table <- data.frame(
  Model    = c("Random Selection (Baseline)","Logistic Regression", "Decision Tree", "Random Forest"),
  Top_N    = rep(top_n_val, 4),
  Renewals = c(random_renewals, prof_logit$renewals, prof_tree$renewals, prof_rf$renewals),
  Cost     = rep(top_n_val * cost_per_contact, 4),
  Revenue  = c(random_renewals, prof_logit$renewals, prof_tree$renewals, prof_rf$renewals) * revenue_per_renewal,
  Profit   = c(random_profit, prof_logit$profit, prof_tree$profit, prof_rf$profit)
)

kable(profit_table,
      col.names = c("Model", "Customers Targeted", "Renewals", "Cost (£)", "Revenue (£)", "Profit (£)"),
      caption   = paste("Profit Simulation on Validation Set (Top", top_n_val, "customers)")) %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(1, italic = TRUE, background = "#F5F5F5") %>%
  row_spec(which.max(profit_table$Profit), bold = TRUE, background = "#D5F5E3")
```

## Selected Model: Random Forest

**Random Forest is selected as the final model.** It consistently achieves the highest AUC and  an ideal profit value across the validation set. Random Forest is a method that reduces variance through bootstrap aggregation and random feature selection, making it more robust to over-fitting than a single decision tree while also remaining more interpretable than gradient boosting methods. In a business context where false negatives (missing likely renewers) are costly, the high sensitivity of the Random Forest model is particularly valuable and makes it a suitable model for the campaign.

---
