---
title: "BU52710 Predictive Analytics – Final Submission"
author: 'Andrew Wilson Student ID: 2417549'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

# Introduction

The locked ten predictor variables and model type outlined from the checkpoint are carried
forward and remaind unchanged. The script covers: model re-training on the original data, full
test-set evaluation (ROC curve, lift chart, profit curve), and business recommendations
across three strategies.

---

# Step 1: Re-load Libraries and Reproduce Training Split

```{r load-libraries}
library(tidyverse)
library(caret)
library(pROC)
library(rpart)
library(rpart.plot)
library(randomForest)
library(car)       # for VIF
library(knitr)
library(kableExtra)
library(ggplot2)
library(corrplot)
library(gridExtra)
```

```{r load-training-data}
# Load the original 8,000-customer dataset (training + validation)
data_train <- read.csv("dataset_2417549_train.csv", stringsAsFactors = TRUE)
data_train$Renewal <- factor(data_train$Renewal, levels = c("No", "Yes"))

# Quick overview
dim(data_train)
str(data_train)

# Reproduce the exact 70/30 stratified split from the checkpoint
set.seed(2417549)
split_index     <- createDataPartition(data_train$Renewal, p = 0.70, list = FALSE)
train_data      <- data_train[split_index, ]
validation_data <- data_train[-split_index, ]

cat("Training set:   ", nrow(train_data),      "rows\n")
cat("Validation set: ", nrow(validation_data), "rows\n")
cat("Train renewal:  ", round(mean(train_data$Renewal == "Yes") * 100, 2), "%\n")
cat("Val renewal:    ", round(mean(validation_data$Renewal == "Yes") * 100, 2), "%\n")

# Verify the class balance is preserved
train_rate <- round(prop.table(table(train_data$Renewal)) * 100, 2)
val_rate   <- round(prop.table(table(validation_data$Renewal)) * 100, 2)

balance_check <- data.frame(
  Set        = c("Training", "Validation"),
  No_pct     = c(train_rate["No"],  val_rate["No"]),
  Yes_pct    = c(train_rate["Yes"], val_rate["Yes"])
)
kable(balance_check,
      col.names = c("Set", "% Non-Renewal", "% Renewal"),
      caption   = "Renewal Rate Balance Across Splits",
      digits    = 2) %>%
  kable_styling(full_width = FALSE)
```

```{r variable-selection}
# LOCKED predictor variables from checkpoint – cannot be changed
selected_vars <- c("PoundsPerIssue", "Income", "Child6.12",
                   "TotalAmountPaid", "MonthsSince1stOrder", "TotalPaidOrders",
                   "Child13.18", "PaidCashMagazines", "HouseholdSize", "HomeValue")

train_sel <- train_data      %>% select(all_of(selected_vars), Renewal)
val_sel   <- validation_data %>% select(all_of(selected_vars), Renewal)
```

---

# Step 2: Re-train Locked Models on Original Training Data

Checkpoint model is re-trained on the same training set for continuity.
The locked final model is the **Random Forest**.

```{r retrain-rf}
set.seed(2417549)
# Model: Random Forest – LOCKED final model
rf_model <- randomForest(Renewal ~ ., data = train_sel,
                         ntree = 500, mtry = 3, importance = TRUE)

# Validation-set predictions (continuity check against checkpoint)
rf_probs_val <- predict(rf_model, newdata = val_sel, type = "prob")[, "Yes"]
rf_pred_val  <- predict(rf_model, newdata = val_sel, type = "class")
cm_rf_val    <- confusionMatrix(rf_pred_val, val_sel$Renewal, positive = "Yes")
roc_rf_val   <- roc(as.numeric(val_sel$Renewal) - 1, rf_probs_val)

cat("Validation AUC – Random Forest:", round(auc(roc_rf_val), 4), "\n")
```

---

# Step 3: Load Test Data and Generate Predictions

```{r load-test-data}
# Load the released 5,000-customer test dataset
test_data <- read.csv("dataset_2417549_final.csv", stringsAsFactors = TRUE)
test_data$Renewal <- factor(test_data$Renewal, levels = c("No", "Yes"))

cat("Test set rows:", nrow(test_data), "\n")

# Verify test set renewal distribution
test_renewal_summary <- test_data %>%
  count(Renewal) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2))

kable(test_renewal_summary,
      col.names = c("Renewal", "Count", "Percentage (%)"),
      caption   = "Renewal Distribution – Test Set") %>%
  kable_styling(full_width = FALSE)

# Subset to the locked 10 predictors
test_sel <- test_data %>% select(all_of(selected_vars), Renewal)
```

```{r test-predictions}
# Apply the locked Random Forest to the full test set
test_probs <- predict(rf_model, newdata = test_sel, type = "prob")[, "Yes"]
test_pred  <- predict(rf_model, newdata = test_sel, type = "class")

# Build the ranked predictions data frame
predictions_df <- data.frame(
  customer_id           = test_data$CustomerID,
  predicted_probability = round(test_probs, 6)
) %>%
  arrange(desc(predicted_probability)) %>%
  mutate(rank = row_number())

# Write submission CSV
write.csv(predictions_df, "2417549_Predictions.csv", row.names = FALSE)
cat("Predictions saved: 2417549_predictions.csv\n")
cat("Total scored:      ", nrow(predictions_df), "\n")
```

---

# Step 4: Test Set Evaluation

## 4.1 Confusion Matrix and Core Metrics

```{r test-confusion}
cm_test  <- confusionMatrix(test_pred, test_sel$Renewal, positive = "Yes")
roc_test <- roc(as.numeric(test_sel$Renewal) - 1, test_probs)
print(cm_test)
cat("\nTest AUC:", round(auc(roc_test), 4), "\n")
```

## 4.2 Validation vs Test Performance Comparison

```{r val-vs-test-table}
perf_comparison <- data.frame(
  Dataset     = c("Validation Set", "Test Set"),
  Accuracy    = round(c(cm_rf_val$overall["Accuracy"],
                        cm_test$overall["Accuracy"]), 4),
  Sensitivity = round(c(cm_rf_val$byClass["Sensitivity"],
                        cm_test$byClass["Sensitivity"]), 4),
  Specificity = round(c(cm_rf_val$byClass["Specificity"],
                        cm_test$byClass["Specificity"]), 4),
  Precision   = round(c(cm_rf_val$byClass["Pos Pred Value"],
                        cm_test$byClass["Pos Pred Value"]), 4),
  AUC         = round(c(auc(roc_rf_val), auc(roc_test)), 4)
)

kable(perf_comparison,
      col.names = c("Dataset", "Accuracy", "Sensitivity",
                    "Specificity", "Precision", "AUC"),
      caption   = "Random Forest Performance: Validation vs Test Set") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(2, bold = TRUE, background = "#D5F5E3")
```

## 4.3 ROC Curve

```{r roc-test}
plot(roc_test, col = "#1E3A8A", lwd = 2.5,
     main = "ROC Curve – Random Forest (Test vs Validation Set)")
plot(roc_rf_val, col = "#A0BDD8", lwd = 2, lty = 2, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey50")
legend("bottomright",
       legend = c(paste("Test Set (AUC =",       round(auc(roc_test),   3), ")"),
                  paste("Validation Set (AUC =",  round(auc(roc_rf_val), 3), ")")),
       col    = c("#1E3A8A", "#A0BDD8"),
       lwd    = c(2.5, 2))
```

## 4.4 Lift Chart

```{r lift-chart}
# Build lift chart data
lift_df <- data.frame(
  prob   = test_probs,
  actual = as.numeric(test_sel$Renewal == "Yes")
) %>%
  arrange(desc(prob)) %>%
  mutate(
    cum_n         = row_number(),
    cum_renewals  = cumsum(actual),
    pct_customers = cum_n / n() * 100,
    lift          = (cum_renewals / cum_n) / mean(actual)
  )

lift_at_10 <- lift_df %>%
  filter(pct_customers <= 10) %>%
  slice_tail(n = 1) %>%
  pull(lift)

cat("Lift @ top 10%:", round(lift_at_10, 2), "\n")

ggplot(lift_df, aes(x = pct_customers, y = lift)) +
  geom_line(colour = "#1E3A8A", linewidth = 1.1) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 10, linetype = "dashed",
             colour = "#C0392B", linewidth = 0.8) +
  annotate("text", x = 11.5, y = max(lift_df$lift) * 0.9,
           label = paste0("Top 10%\nLift = ", round(lift_at_10, 1), "x"),
           colour = "#C0392B", size = 3.2, hjust = 0) +
  labs(title    = "Cumulative Lift Chart – Random Forest (Test Set)",
       subtitle = "Horizontal dashed line = random selection (Lift = 1.0)",
       x        = "Percentage of Customers Contacted (%)",
       y        = "Lift") +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  theme_minimal() +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 10, colour = "grey40"),
    panel.grid.minor = element_blank()
  )
```

## 4.5 Profit Curve

```{r profit-curve}
cost_per_contact    <- 40
revenue_per_renewal <- 140

profit_df <- lift_df %>%
  mutate(profit = cum_renewals * revenue_per_renewal - cum_n * cost_per_contact)

opt_row <- profit_df %>% slice_max(profit, n = 1, with_ties = FALSE)

ggplot(profit_df, aes(x = cum_n, y = profit)) +
  geom_line(colour = "#1E3A8A", linewidth = 1.1) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_vline(xintercept = 500, linetype = "dashed",
             colour = "#C0392B", linewidth = 0.8) +
  annotate("text", x = 515, y = min(profit_df$profit) * 0.5,
           label = "Budget cap & optimal\n(n = 500)", colour = "#C0392B",
           size = 3, hjust = 0) +
  scale_y_continuous(labels = function(x) paste0("\u00A3", format(x, big.mark = ",", scientific = FALSE))) +
  labs(title    = "Profit Curve - Random Forest (Test Set)",
       subtitle = paste0("Profit at every targeting threshold; parameters: \u00A3", "40 cost, \u00A3", "140 revenue"),
       x        = "Number of Customers Contacted",
       y        = paste0("Estimated Profit (", "\u00A3", ")")) +
  theme_minimal() +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 10, colour = "grey40"),
    panel.grid.minor = element_blank()
  )

cat("Budget-cap profit (n = 500):", paste0("\u00A3", format(profit_df$profit[500], big.mark = ",")), "\n")
cat("Optimal profit    (n =", opt_row$cum_n, "):", paste0("\u00A3", format(opt_row$profit, big.mark = ",")), "\n")
```

---

# Step 5: Targeting Strategies

## 5.1 Strategy 1 – Full Budget (Top 500 Customers)

```{r top-500-strategy}
cost_per_contact    <- 40
revenue_per_renewal <- 140

test_ranked <- predictions_df %>%
  left_join(test_data %>% select(CustomerID, Renewal),
            by = c("customer_id" = "CustomerID"))

top500        <- test_ranked %>% slice_head(n = 500)
renewals_500  <- sum(top500$Renewal == "Yes")
resp_rate_500 <- round(renewals_500 / 500 * 100, 2)
cost_500      <- 500 * cost_per_contact
revenue_500   <- renewals_500 * revenue_per_renewal
profit_500    <- revenue_500 - cost_500

cat("Customers targeted:", 500,           "\n")
cat("Renewals:          ", renewals_500,  "\n")
cat("Response rate:     ", resp_rate_500, "%\n")
cat("Cost:              \u00A3", format(cost_500,    big.mark = ","), "\n")
cat("Revenue:           \u00A3", format(revenue_500, big.mark = ","), "\n")
cat("Profit:            \u00A3", format(profit_500,  big.mark = ","), "\n")
```

## 5.2 Strategy 2 – Optimal Targeting

```{r optimal-strategy}
opt_strat <- test_ranked %>%
  mutate(
    is_renewal   = as.numeric(Renewal == "Yes"),
    cum_renewals = cumsum(is_renewal),
    profit_at_n  = cum_renewals * revenue_per_renewal - rank * cost_per_contact
  ) %>%
  slice_max(profit_at_n, n = 1, with_ties = FALSE)

opt_n        <- as.numeric(opt_strat$rank)
opt_renewals <- as.numeric(opt_strat$cum_renewals)
opt_profit   <- as.numeric(opt_strat$profit_at_n)
opt_cost     <- opt_n * cost_per_contact
opt_revenue  <- opt_renewals * revenue_per_renewal
opt_resp     <- round(opt_renewals / opt_n * 100, 2)

cat("Optimal customers: ", opt_n,        "\n")
cat("Renewals:          ", opt_renewals, "\n")
cat("Response rate:     ", opt_resp,     "%\n")
cat("Cost:              \u00A3", format(opt_cost,    big.mark = ","), "\n")
cat("Revenue:           \u00A3", format(opt_revenue, big.mark = ","), "\n")
cat("Profit:            \u00A3", format(opt_profit,  big.mark = ","), "\n")
```

## 5.3 Strategy Comparison Table

```{r strategy-comparison}
test_renewal_rate <- mean(test_data$Renewal == "Yes")
rand_renewals     <- round(500 * test_renewal_rate)
rand_cost         <- 500 * cost_per_contact
rand_revenue      <- rand_renewals * revenue_per_renewal
rand_profit       <- rand_revenue - rand_cost

strategy_tbl <- data.frame(
  Strategy  = c("Random Selection (Baseline)",
                "Model - Top 500 (Full Budget)",
                "Model - Optimal Threshold"),
  Customers = c(500, 500, opt_n),
  Cost      = c(rand_cost,     cost_500,     opt_cost),
  Renewals  = c(rand_renewals, renewals_500, opt_renewals),
  Revenue   = c(rand_revenue,  revenue_500,  opt_revenue),
  Profit    = c(rand_profit,   profit_500,   opt_profit)
)
kable(strategy_tbl,
      col.names = c("Strategy", "Customers", "Cost (\u00A3)",
                    "Renewals", "Revenue (\u00A3)", "Profit (\u00A3)"),
      caption   = "Business Strategy Comparison - Test Set",
      digits    = 0,
      format.args = list(big.mark = ",")) %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(1, italic = TRUE, background = "#F5F5F5") %>%
  row_spec(3, bold = TRUE, background = "#D5F5E3")
```

---